---
title: RPE Cells
---

# Needed packages

```{r, message=FALSE}
require(tidyverse)
require(igraph)
require(ggplot2)
require(vioplot)
```

# Needed data

All the needed data are available [here](https://github.com/ocbe-uio/CellMAPtracer/tree/master/Data/RPE%20tracking%20data)

The notebook consists of five sections

# Section 1: Studying Dividing Daughter Cells

## Loading the data

First we create a `path` string that stores the location of the files:

```{r set-path}
path <- "../Data/RPE tracking data/" # You may need to change this
```

Now we can properly read the files.


```{r}
df <- read.csv(file.path(path, "PIP-FUCCI Dividing Daughter Cells.csv"))
head(df)
dim(df)
```

## Computing the doubling time of RPE cells

```{r}
TrajectoryTime <- df[, 9]/60
sd(TrajectoryTime)
summary(TrajectoryTime)
```

## Computing the mode

```{r}
getmode <- function(v) {
	uniqv <- unique(v)
	uniqv[which.max(tabulate(match(v, uniqv)))]
}
result <- getmode(TrajectoryTime)
print(result)
```

## Plotting the density of the doubling time

```{r}
x <- TrajectoryTime
hx5 <- TrajectoryTime
dens <- density(hx5, cut=10)
n <- length(dens$y)
dx <- mean(diff(dens$x))                  # Typical spacing in x
y.unit <- sum(dens$y) * dx                # Check: this should integrate to 1
dx <- dx  /  y.unit                       # Make a minor adjustment
x.mean <- sum(dens$y * dens$x) * dx
y.mean <- dens$y[length(dens$x[dens$x < x.mean])]
x.mode <- dens$x[i.mode <- which.max(dens$y)]
y.mode <- dens$y[i.mode]
y.cs <- cumsum(dens$y)
x.med <- dens$x[i.med <- length(y.cs[2 * y.cs <= y.cs[n]])]
y.med <- dens$y[i.med]
```

## Plotting the density and the statistics.

```{r}
plot(
	dens, type="l", col="black", lwd=0.1, xlim=c(0, 25), cex=1.5, cex.lab=1.3,
	cex.axis=1.3, xlab="The doubling time of hTERT-immortalized RPE cells (h)"
)
polygon(dens, col="black")
temp <- mapply(
	function(x, y, c) lines(c(x, x), c(0, y), lwd=2, col=c, las=1),
	c(x.mean, x.med, x.mode),
	c(y.mean, y.med, y.mode),
	c("orange", "Green", "Red")
)
legend(
	14, 0.24, c("Doubling Time Density", "Mode", "Median", "Mean"),
	col=c("black", "Red", "green", "orange"), text.col="black", lty=1, lwd=2,
	cex=1, merge=TRUE
)
```

## Characterizing the trajectory movement of a population of Dividing Daughter cells

### Directionality

```{r}
Directionality <- df[, 10]
Group <- c(rep(1, length(Directionality)))
Da <- data.frame(Directionality, Group)

#  Add mean and standard deviation within violin plot
data_summary <- function(x) {
	m <- mean(x)
	ymin <- m - sd(x)
	ymax <- m + sd(x)
	return(c(y=m, ymin=ymin, ymax=ymax))
}

p <- ggplot(Da, aes(x=Group, y=Directionality)) +
	geom_violin(trim=FALSE, fill="lightgreen") +
	geom_point(size=2.5, color="black")
p + stat_summary(fun.data=data_summary, geom="pointrange", color="red", size=1)
```

### Speed

```{r}
Average_Speed <- df[, 11] * 60
Group <- c(rep(1, length(Average_Speed)))
Da <- data.frame(Average_Speed, Group)
data_summary <- function(x) {
	m <- mean(x)
	ymin <- m - sd(x)
	ymax <- m + sd(x)
	return(c(y=m, ymin=ymin, ymax=ymax))
}
p <- ggplot(Da, aes(x=Group, y=Average_Speed)) +
	geom_violin(trim=FALSE, fill="yellow") +
	geom_point(size=2.5, color="black")
p + stat_summary(fun.data=data_summary, geom="pointrange", color="red", size=1)
```

### Correlation Analysis

#### Finding the correlation between doubling time and Total Distance

```{r}
DT <- df[, 9] / 60    # doubling time
f <- df[, 7]          # Total Distance
ff <- as.numeric(gsub(", ", ".", f))
plot(
	DT, f, xlab="Doubling time (h)", ylab="Total Distance (um)", pch=16, las=0,
	cex=1.5, cex.lab=1.3, cex.axis=1.3
)
c <- cor.test(~ DT+ ff, method="pearson", exact=FALSE)
c
reg <- lm(ff ~ DT)
cc <- unlist(c[4])
ccPV <- round(cc, digits=2)
abline(reg, untf=TRUE, col="red", lwd=2)
text(18, 1875, "r = 0.62", cex=1.5)
text(19, 1750, "P < 0.001", cex=1.5)
```

#### Finding the correlation between doubling time and Directionality

```{r}
DT <- df[, 9] / 60    # doubling time
f <- df[, 10]         # Directionality
plot(
	DT, f, xlab="Doubling time (h)", ylab="Directionality", pch=16, las=1,
	ylim=c(0, .65)
)
c <- cor.test( ~ DT + f, method="pearson", exact=FALSE)
c
reg <- lm(f ~ DT)
cc <- unlist(c[4])
ccPV <- round(cc, digits=2)
abline(reg, untf=TRUE, col="red", lwd=2)
text(19, 0.65, "r = -0.14", cex=1)
text(19.4, 0.6, "P = 0.057", cex=1)
```

#### Finding the correlation between doubling time and Average Speed

```{r}
DT <- df[, 9] / 60    # doubling time
f <- df[, 11] * 60    # Average Speed
plot(
	DT, f, xlab="Doubling time (h)", ylab="Average Speed (um / h)", pch=16,
	las=1, ylim=c(10, 60)
)
c <- cor.test( ~ DT+ f, method="pearson", exact=FALSE)
c
reg <- lm(f ~ DT)
cc <- unlist(c[4])
ccPV <- round(cc, digits=2)
abline(reg, untf=TRUE, col="red", lwd=2)
text(40.1, 60, "r = -0.12", cex=1)
text(40, 57, "P = 0.1", cex=1)
```

# Section 2: Plotting the lineage tree

## Loading the data

```{r}
df <- read.csv(file.path(path, "PIP-FUCCI All Cells_Results.csv"))
head(df)
dim(df)
```

## Lineage tree of each cell and its descendants with edge length linked to time till cell division

```{r}
DF <- df
ID <- DF[, 3]
SS <- sub("*\\..*", "", unlist(ID))

DF <- data.frame(ID, SS)
DFF <- split(DF[, 1], DF[, 2])
DFFF <- DFF
for (j in 1:length(DFF)) {
	if (length(DFF[[j]]) < 2) {
		DFFF[j]<-NULL
	}
}
length(DFFF)
DFF <- DFFF
TT <- c() # to be used for showing the edge length
for (i in 1:length(DFFF)) {
	TT <- c(TT, df[which(df[, 3] == DF[i, 1]), 6])
}
for (j in 1:length(DFF)) {
	parents <- c()
	Offspring <- c()
	col <- c()
	test <- sub("*\\..d*$", "", unlist(DFF[[j]]))
	for (i in 1:length(test)) {
		if (!test[i] == DFF[[j]][i]) {
			parents <- c(parents, test[i])
			Offspring <- c(Offspring, DFF[[j]][i])
		}
	}
	d <- tibble(Offspring, parents)
	d2 <- data.frame(from=d$parents, to=d$Offspring)

	g <- graph_from_data_frame(d2)
	ordered.vertices <- get.data.frame(g, what="vertices")
	for (n in 1:length(parents)) {
		if (Offspring[n] %in% parents) {
			col <- c(col, Offspring[n])
		}
	}

	COLOR <- c(rep("gray", length(ordered.vertices[, 1])))
	for (i in 1:length(ordered.vertices[, 1])) {
		if (ordered.vertices[i, 1] %in% col) {
			COLOR[i] <- "lightgreen"
		}
	}

	TTT <- TT / TT[which.max(TT)]
	E(g)$weight <- TTT

	set.seed(100)
	test.layout <- layout_(g, with_dh(weight.edge.lengths=TTT))
	test.layout[, 1] <- test.layout[, 1] * 2
	test.layout[, 2] <- test.layout[, 2] * 3.5
}
```

Plots can be generated and saved to the user's working directory with

```{r eval=FALSE}
pdf(paste0(j, ".plot.pdf"))
par(mar=c(0, 0,0, 0) + .1)
plot(
	g, layout=test.layout, edge.arrow.size=0.64, edge.arrow.width=2,
	edge.width=3.5, vertex.shape="circle", vertex.size=22,
	edge.color="black", vertex.color=adjustcolor(COLOR, alpha.f=.999),
	vertex.label.cex=0.65, vertex.label.font=2, vertex.label.color="black"
)
dev.off()
```

## Plotting the lineage tree of a particular cell and its descendants

```{r}
DF <- df
ID <- DF[, 3]
SS <- sub("*\\..*", "", unlist(ID))

DF <- data.frame(ID, SS)
DFF <- split(DF[, 1], DF[, 2])
DFFF <- DFF
for (j in 1:length(DFF)) {
	if (length(DFF[[j]]) < 2) {
		DFFF[j] <- NULL
	}
}
length(DFFF)
DFF <- DFFF
TT <- c() # to be used for showing the edge length
for (i in 1:length(DFFF)) {
	TT <- c(TT, df[which(df[, 3] == DF[i, 1]), 6])
}
j <- 10 # this changes the target cell and generat different lineage tree
parents <- c()
Offspring <- c()
col <- c()
test <- sub("*\\..d*$", "", unlist(DFF[[j]]))
for (i in 1:length(test)) {
	if (!test[i] == DFF[[j]][i]) {
		parents <- c(parents, test[i])
		Offspring <- c(Offspring, DFF[[j]][i])
	}
}
d <- tibble(Offspring, parents)
d2 <- data.frame(from=d$parents, to=d$Offspring)
g <- graph_from_data_frame(d2)
ordered.vertices <- get.data.frame(g, what="vertices")
for (n in 1:length(parents)) {
	if (Offspring[n] %in% parents) {
		col <- c(col, Offspring[n])
	}
}
COLOR <- c(rep("gray", length(ordered.vertices[, 1])))
for (i in 1:length(ordered.vertices[, 1])) {
	if (ordered.vertices[i, 1] %in% col) {
		COLOR[i] <- "lightgreen"
	}
}
TTT <- TT / TT[which.max(TT)]
E(g)$weight <- TTT
set.seed(100)
test.layout <- layout_(g, with_dh(weight.edge.lengths=TTT))
test.layout[, 1] <- test.layout[, 1] * 2
test.layout[, 2] <- test.layout[, 2] * 3.5
par(mar=c(0, 0,0, 0) + .1)
plot(
	g, layout=test.layout, edge.arrow.size=0.64, edge.arrow.width=2,
	edge.width=3.5, vertex.shape="circle", vertex.size=22, edge.color="black",
	vertex.color=adjustcolor(COLOR, alpha.f=.999), vertex.label.cex=0.65,
	vertex.label.font=2, vertex.label.color="black"
)
```

# Section 3: Generation plot for all the data

## Loading the data after converting each file into csv

```
df <- read.csv(file.path(path, "PIP-FUCCI All Cells_Results.csv"))
head(df)
dim(df)
mothers <- c()
first <- c()
second <- c()
third <- c()
fourth <- c()
for (i in 1:length(df[, 1])) {
	st <- strsplit(df[i, 3], regex("\\.", multiline=TRUE))
	L <- length(st[[1]]) - 1
	if (L == 0) {mothers=c(mothers, L)}
	if (L == 1) {first=c(first, L)}
	if (L == 2) {second=c(second, L)}
	if (L == 3) {third=c(third, L)}
	if (L == 4) {fourth=c(fourth, L)}
}
generations <- c(length(third), length(second), length(first), length(mothers))
barplot(
	generations, horiz=TRUE, col=c("bisque1", "bisque2", "bisque3", "bisque4"),
	xlab="Number of cells", xlim=c(0, 100)
)
box()
```

# Section 4: Heterogeneity between daughter cells

## Loading the data after converting each file into csv

```{r}
dcells <- read.csv(file.path(path, "PIP-FUCCI Daughter Cells_Results.csv"))
dim(dcells)
head(dcells)

firstDC <- c() # selecting first DCell
num <- c(1:length(dcells[, 1]))
for (k in num) {
	if(!(k %% 2) == 0) {
		firstDC <- c(firstDC, k)
	}
}

sdDiff <- 15  # no trajectory time difference larger than 1.3 hours between the two daughter cells
minStep <- 36 # minimum trajectory time of 3 hours

selectedrows <- c()
for (i in firstDC) {
	condition1 <- abs(dcells[i, 6] - dcells[i + 1, 6]) <= sdDiff
	condition2 <- dcells[i, 6] >= minStep
	condition3 <- dcells[i + 1, 6] >= minStep
	if (condition1 & condition2 & condition3) {
		selectedrows <- c(selectedrows, i)
	}
}

Fselectedrows <- c(selectedrows, (selectedrows + 1)) # including the other cell
Fselectedrows <- sort(Fselectedrows)
Fdcells <- dcells[Fselectedrows, ]
diffDC <- dcells[selectedrows, ]
for (i in selectedrows) {
	diffDC[which(selectedrows %in% i), 5:11] <- abs(
		dcells[i, 5:11] - dcells[i + 1, 5:11]
	)
}
length(diffDC[, 1])         # pairs of daughter cells included in the analysis
summary(diffDC[, 11] * 60)  # speed difference
summary(diffDC[, 7])        # total distance difference
summary(diffDC[, 10] * 100) # directionality difference (%)
summary(diffDC[, 8])        # displacement difference
par(mfrow=c(1, 4))
boxplot(
	diffDC[, 11] * 60, ylab="Average Speed difference (um / h)", cex.lab=1.2,
	cex.axis=1.2, col="yellow", ylim=c(0, 10)
)
boxplot(
	diffDC[, 10] * 100, ylab="Directionality difference (%)", cex.lab=1.2,
	cex.axis=1.2, col="lightgreen", ylim=c(0, 40)
)
boxplot(
	diffDC[, 7], ylab="Total Distance difference (um)", cex.lab=1.2,
	cex.axis=1.2, col="lightblue", ylim=c(0, 100)
)
boxplot(
	diffDC[, 8], ylab="Displacement difference (um)", cex.lab=1.2,
	cex.axis=1.2, col="blue", ylim=c(0, 100)
)
```

## Computing the Time Diffrence between divisions of RPE daughter cells

```{r}
TimeDifferences <- c()
for (i in firstDC) {
	if (dcells[i, 4] == TRUE & dcells[i + 1, 4] == TRUE) {
		h <- abs(dcells[i, 9] - dcells[i + 1, 9])
		TimeDifferences <- c(TimeDifferences, h)
	}
}
length(TimeDifferences) # pairs of daughter cells included in the analysis

summary(TimeDifferences / 60)
sd(TimeDifferences / 60)
```

## Computing the mode

```{r}
getmode <- function(v) {
	uniqv <- unique(v)
	uniqv[which.max(tabulate(match(v, uniqv)))]
}
result <- getmode(TimeDifferences)
print(result)
```

## Boxplot

```
boxplot(
	TimeDifferences / 60,
	ylab="Time difference between the division of daughter cells (h)",
	cex.lab=1.2, cex.axis=1.2, col="gray", ylim=c(0, 15), las=1
)
```

## Violin plot

```{r}
TD <- TimeDifferences / 60
Group <- c(rep(1, length(TD)))
Da <- data.frame(TD, Group)
data_summary <- function(x) {
	m <- mean(x)
	ymin <- m - sd(x)
	ymax <- m + sd(x)
	return(c(y=m, ymin=ymin, ymax=ymax))
}
p <- ggplot(Da, aes(x=Group, y=TD)) +
	geom_violin(trim=FALSE, fill="gray") +
	geom_point(size=2.5, color="black")
p + stat_summary(fun.data=data_summary, geom="pointrange", color="red", size=1)
```

# Section 5: Changes across cell cycle phases

## Loading the RPE FUCCI cell-cycle dividing daughter cells

```{r}
dff <- read.csv(
	file.path(path, "RPE_FUCCI_cell-cycle_dividing_daughter_cells.csv")
)
dim(dff)
head(dff)
```

## Setting a value for the pixel size based on the tiff files

```{r}
pixel_size <- 0.65
```

## Removing cells with no cell cycle phases

```{r}
Cells <- subset(dff, dff[, 9] != "") # removing rows with NA
dim(Cells)
row.has.na <- apply(Cells, 1, function(x) {any(is.na(x))}) # check if rows NAs
Cells <- Cells[!row.has.na, ]
dim(Cells)
Cells [, 4] <- Cells [, 4] * pixel_size
Cells [, 5] <- Cells [, 5] * pixel_size
SP1 <- split(Cells, Cells[, 2]) # splitting cells based on their ID
length(SP1) # getting the number of dividing daughter cells
for (u in 1:length(SP1)) {
	SP1[[u]] <- split(SP1[[u]], SP1[[u]][, 9]) # split each cell based on phases
}
```

## Computing the average length of each phase as a % of the total length of the cell cycle

```{r}
G1 <- c()
G2 <- c()
S <- c()
for (p in 1:length(SP1)) {
	denominator <- sum(
		length(SP1[[p]][[1]][, 1]),
		length(SP1[[p]][[2]][, 1]),
		length(SP1[[p]][[3]][, 1])
	)
	g1 <- length(SP1[[p]][[1]][, 1]) / denominator
	g2 <- length(SP1[[p]][[2]][, 1]) / denominator
	s <- length(SP1[[p]][[3]][, 1]) / denominator
	G1 <- c(G1, g1)
	G2 <- c(G2, g2)
	S <- c(S, s)
}
mG1 <- round(mean(G1) * 100)
sdG1 <- round(sd(G1) * 100)
mS <- round(mean(S) * 100)
sdS <- round(sd(S) * 100)
mG2 <- round(mean(G2) * 100)
sdG2 <- round(sd(G2) * 100)
```

## Plotting the Phase average length (%)

### Barplot

```
yy <- c(mG1, mS, mG2)
std1 <- c(sdG1, sdS, sdG2)
pp <- barplot(
	yy, ylim=c(0, 100), xlab="Cell cycle phases",
	ylab="Phase average length (%)", las=1, names.arg=c("G1", "S", "G2"),
	col=c("khaki2", "khaki3", "khaki4")
)
segments(pp, yy-std1, pp, yy + std1)
segments(pp - 0.1, yy - std1, pp + 0.1, yy - std1)
segments(pp - 0.1, yy + std1, pp + 0.1, yy + std1)
box()
```

## Pairwise comparisons using Wilcoxon rank sum test with continuity correction

```{r}
y <- c(G2, S, G1)
Group <- factor(
	c(rep("G1", 43), rep("S", 43), rep("G2", 43)), levels=c("G1", "S", "G2")
)
Group <- factor(Group, levels=c("G1", "S" ,"G2")) # define display order
df <- data.frame(y, Group)
kruskal.test(y ~ Group, data=df)
pairwise.wilcox.test(y, Group, p.adj="none", exact=FALSE)
```
## Computing for each step the distance and Instantaneous Speed

```{r}
TimeInterval <- 5
for (p in 1:length(SP1)) {
	for (v in 1:3) {
		M <- SP1[[p]][[v]]
		MM <- length(M[, 1])
		res <- t(
			sapply(
				1:MM,
				function(i) {
					# creating values for dx
					SP1[[p]][[v]][i, 10] <- (SP1[[p]][[v]][i + 1, 4]) -
						(SP1[[p]][[v]][i, 4])
					SP1[[p]][[v]][, 10][is.na(SP1[[p]][[v]][, 10])] <- 0
					# creating values for dy
					SP1[[p]][[v]][i, 11] <- (SP1[[p]][[v]][i + 1, 5]) -
						(SP1[[p]][[v]][i, 5])
					# to remove NA and replace it with 0
					SP1[[p]][[v]][, 11][is.na(SP1[[p]][[v]][, 11])] <- 0
					# creating values for dis
					SP1[[p]][[v]][i, 12] <- sqrt((SP1[[p]][[v]][i, 10]) ^ 2 +
						(SP1[[p]][[v]][i, 11]) ^ 2)
					# creating values for Square Speed
					SP1[[p]][[v]][i, 13] <-
						((SP1[[p]][[v]][i, 12]) / TimeInterval) * 60
					return(SP1[[p]][[v]][i, c(10:13)])
 				}
			)
		)
		SP1[[p]][[v]][1:MM, c(10:13)] <- as.data.frame(res)
		SP1[[p]][[v]][, c(10:13)] <- lapply(
			SP1[[p]][[v]][, c(10:13)], as.numeric
		)
		colnames(SP1[[p]][[v]]) <- c(
			"CellID", "CellName", "ImageID", "X", "Y", "R", "G", "B", "phase",
			"dX", "dY", "Dis", "Speed"
		)
	}
}
```

## Average Instantaneous Speed

```{r}
speedG1 <- c()
speedG2 <- c()
speedS <- c()
for (p in 1:length(SP1)) {
	sG1 <- mean(SP1[[p]][[1]][, 13])
	speedG1 <- c(speedG1, sG1)
	sG2 <- mean(SP1[[p]][[2]][, 13])
	speedG2 <- c(speedG2, sG2)
	sS <- mean(SP1[[p]][[3]][, 13])
	speedS <- c(speedS, sS)
}
summary(speedG1)
sd(speedG1)
summary(speedS)
sd(speedS)
summary(speedG2)
sd(speedG2)
```

## Plotting the Average Instantaneous Speeds across cell cycle

```{r}
x1 <- speedG1
x2 <- speedS
x3 <- speedG2
```

### Violin Plot

```{r}
plot(
	1, 1, xlim=c(0, 4), ylim=c(0, 60), type='n', xlab='Cell Cycle Phases',
	ylab='', xaxt='n', las=1
)
vioplot(x1, at=1, add=TRUE, col="khaki2")
vioplot(x2, at=2, add=TRUE, col="khaki3")
vioplot(x3, at=3, add=TRUE, col="khaki4")
axis(1, at=c(1, 2,3), labels=c('G1', "S", "G2"))
axis(2, at =30, pos=-0.45, tck=0, labels='Average Instantaneous Speeds (um/h)')
```

### Boxplot

```{r}
speed <- c(speedG1, speedS, speedG2)
phases <- factor(
	c(rep("G1", 43), rep("S", 43), rep("G2", 43)), levels=c("G1", "S",  "G2")
)
myDF <- data.frame(speed, phases)
boxplot(
	speed ~ phases, data=myDF, las=1, col=c("khaki2", "khaki3", "khaki4"),
	ylab="Average Instantaneous Speeds (um/h)", xlab="Cell Cycle Phases"
)
```

### Barplot

```{r}
msG1 <- mean(speedG1)
sdG1 <- sd(speedG1)
msS <- mean(speedS)
sdS <- sd(speedS)
msG2 <- mean(speedG2)
sdG2 <- sd(speedG2)
yy <- c(msG1, msS, msG2)
std1 <- c(sdG1, sdS, sdG2)
pp <- barplot(
	yy, ylim=c(0, 50), xlab="Cell cycle phases",
	ylab="Average Instantaneous Speeds (um/h)", las=1,
	names.arg=c("G1", "S", "G2"), col=c("khaki2", "khaki3", "khaki4")
)
segments(pp, yy - std1, pp, yy + std1)
segments(pp - 0.1, yy - std1, pp + 0.1, yy - std1)
segments(pp - 0.1, yy + std1, pp + 0.1, yy + std1)
box()
```

## Pairwise comparisons using Wilcoxon rank sum test with continuity correction

```{r}
y <- c(speedG1, speedS, speedG2)
Group <- factor(
	c(rep("G1", 43), rep("S", 43), rep("G2", 43)), levels=c("G1", "S",  "G2")
)
Group <- factor(Group, levels=c("G1", "S" ,"G2")) # order to be displayed
df <- data.frame(y, Group)
kruskal.test(y ~ Group, data=df)
pairwise.wilcox.test(y, Group, p.adj="none", exact=FALSE)
```

## Directionality

### Computing Total Distance and Displacement

```{r}
disG1 <- c()
displacementG1 <- c()
disG2 <- c()
displacementG2 <- c()
disS <- c()
displacementS <- c()
for (p in 1:length(SP1)) {
	sG1 <- sum(SP1[[p]][[1]][, 12])
	disG1 <- c(disG1, sG1)
	dG1 <- sqrt(
		(SP1[[p]][[1]][length(SP1[[p]][[1]][, 1]), 4] - SP1[[p]][[1]][1, 4])^2 +
		(SP1[[p]][[1]][length(SP1[[p]][[1]][, 1]), 5] - SP1[[p]][[1]][1, 5])^2
	)
	displacementG1 <- c(displacementG1, dG1)

	sG2 <- sum(SP1[[p]][[2]][, 12])
	disG2 <- c(disG2, sG2)
	dG2 <- sqrt(
		(SP1[[p]][[2]][length(SP1[[p]][[2]][, 1]), 4] - SP1[[p]][[2]][1, 4])^2 +
		(SP1[[p]][[2]][length(SP1[[p]][[2]][, 1]), 5] - SP1[[p]][[2]][1, 5])^2
	)
	displacementG2 <- c(displacementG2, dG2)

	sS <- sum(SP1[[p]][[3]][, 12])
	disS <- c(disS, sS)
	dS <- sqrt(
		(SP1[[p]][[3]][length(SP1[[p]][[3]][, 1]), 4] - SP1[[p]][[3]][1, 4])^2 +
		(SP1[[p]][[3]][length(SP1[[p]][[3]][, 1]), 5] - SP1[[p]][[3]][1, 5])^2
	)
	displacementS <- c(displacementS, dS)
}

DRg1 <- displacementG1 / disG1
DRg2 <- displacementG2 / disG2
DRs <- displacementS / disS
mean(DRg1)
sd(DRg1)
mean(DRs)
sd(DRg2)
mean(DRg2)
sd(DRg2)
```

### Plotting directionality

#### Boxplot

```{r}
DIS <- c(DRg1, DRs, DRg2)
phases <- factor(
	c(rep("G1", 43), rep("S", 43), rep("G2", 43)), levels=c("G1", "S",  "G2")
)
myDF <- data.frame(DIS, phases)
boxplot(
	DIS ~ phases, data=myDF, las=1, col=c("khaki2", "khaki3", "khaki4"),
	ylab="Directionality Ratio", xlab="Cell Cycle Phases"
)
```

#### Barplot

```{r}
msG1 <- mean(DRg1)
sdG1 <- sd(DRg1)
msS <- mean(DRs)
sdS <- sd(DRs)
msG2 <- mean(DRg2)
sdG2 <- sd(DRg2)
yy <- c(msG1, msS, msG2)
std1 <- c(sdG1, sdS, sdG2)
pp <- barplot(
	yy, ylim=c(0, 0.6), xlab="Cell cycle phases", ylab="Directionality", las=1,
	names.arg=c("G1", "S", "G2"), col=c("khaki2", "khaki3", "khaki4")
)
segments(pp, yy - std1, pp, yy + std1)
segments(pp - 0.1, yy - std1, pp + 0.1, yy - std1)
segments(pp - 0.1, yy + std1, pp + 0.1, yy + std1)
box()
```

## Pairwise comparisons using Wilcoxon rank sum test with continuity correction

```{r}
y <- c(DRg1, DRs, DRg2)
Group <- factor(
	c(rep("G1", 43), rep("S", 43), rep("G2", 43)), levels=c("G1", "S",  "G2")
)
Group <- factor(Group, levels=c("G1", "S" ,"G2")) # order to be displayed
df <- data.frame(y, Group)
kruskal.test(y ~ Group, data=df)
pairwise.wilcox.test(y, Group, p.adj="none", exact=FALSE)
```

## Speed Profiling

```{r}
SP2 <- split(Cells, Cells[, 2]) # splitting each cell based on phases
TimeInterval <- 5
for (p in 1:length(SP2)) {
	M <- SP2[[p]]
	MM <- length(M[, 1])
	res <- t(
		sapply(
			1:MM,
			function(i) {
				# creating values for dx
				SP2[[p]][i, 10] <- (SP2[[p]][i + 1, 4]) - (SP2[[p]][i, 4])
				SP2[[p]][, 10][is.na(SP2[[p]][, 10])] <- 0
				# creating values for dy
				SP2[[p]][i, 11] <- (SP2[[p]][i + 1, 5]) - (SP2[[p]][i, 5])
				# to remove NA and replace it with 0
				SP2[[p]][, 11][is.na(SP2[[p]][, 11])] <- 0
				# creating values for dis
				SP2[[p]][i, 12] <- sqrt(
					(SP2[[p]][i, 10]) ^ 2 + (SP2[[p]][i, 11]) ^ 2
				)
				# creating values for Square Speed
				SP2[[p]][i, 13] <- ((SP2[[p]][i, 12])/TimeInterval) * 60
				return(SP2[[p]][i, c(10:13)])
			}
		)
	)
	SP2[[p]][1:MM, c(10:13)] <- as.data.frame(res)
	SP2[[p]][, c(10:13)] <- lapply(SP2[[p]][, c(10:13)], as.numeric)
	colnames(SP2[[p]]) <- c(
		"CellID", "CellName", "ImageID", "X", "Y", "R", "G", "B", "Phase",
		"dX", "dY", "Dis", "Speed"
	)
}
```

## plotting examples of jump

```{r}
time <- c(1:144)
length(time)
```


```{r}
# An example of a cell with a jump
i <- 1
x <- c(
	1, 6,12, 18, 24, 30, 36, 42, 48, 54, 60, 66, 72, 78, 84, 90, 96, 102, 108,
	114, 120, 126, 132, 138, 144
)
y <- c(
	-24, -23, -22, -21, -20, -19, -18, -17, -16, -15, -14, -13, -12, -11, -10,
	-9, -8, -7, -6, -5, -4, -3, -2, -1, 0
)

x <- c(1, 12, 24, 36, 48, 60, 72, 84, 96, 108, 120, 132, 144)
y <- c(-12, -11, -10, -9, -8, -7, -6, -5, -4, -3, -2, -1, 0)

if (length(SP2[[i]][, 13]) <= 144) {
	missing <- 144 - length(SP2[[i]][, 13])
	P <- c(rep(NA, missing), SP2[[i]][, 13])
	length(P)
	plot(
		time, P,type="l", lwd=3, col="orange", ylim=c(0, 150), xaxt="n",
		xlab="Cell Division Time (h)", ylab="Speed (um/h)", las=1
	)
	axis(1, at=x, labels=y, col.axis="black", cex.axis=0.73, las=1)
} else {
	extra <- 144 - length(SP2[[i]][, 13])
	P <- SP2[[i]][(1 + abs(extra)):length(SP2[[i]][, 13]), 13]
	plot(
		time, P, type="l", lwd=3, col="red", ylim=c(0, 150), xaxt="n",
		xlab="Cell Division Time (h)", ylab="Speed (um/h)", las=1
	)
	axis(1, at=x, labels=y, col.axis="black", las=1)
}

# An example of a cell without a jump
i <- 35
missing <- 144 - length(SP2[[i]][, 13])
P <- c(rep(NA, missing), SP2[[i]][, 13])
lines(time, P, lwd=2.2, col="blue")
add_legend <- function(...) {
	opar <- par(
		fig=c(0, 1, 0, 1), oma=c(0, 0, 0, 0), mar=c(0, 0, 0, 0), new=TRUE
	)
	on.exit(par(opar))
	plot(0, 0, type='n', bty='n', xaxt='n', yaxt='n')
	legend(...)
}
add_legend(
	"top",
	legend=c(
		"Cell ID: C8.2  [A cell without a terminal speed jump]",
		"Cell ID: C10.1  [A cell with a terminal speed jump]"
	),
	col=c("blue", "orange"), horiz=FALSE, bty='n', cex=1, lty=1, lwd=2.4
)
```

##  JUMP

```{r}
SP2 <- split(Cells, Cells[, 2]) # splitting each cell based on phases
TimeInterval <- 5
for (p in 1:length(SP2)) {
	M <- SP2[[p]]
	MM <- length(M[, 1])
	res <- t(
		sapply(
			1:MM,
			function(i) {
				# creating values for dx
				SP2[[p]][i, 10] <- (SP2[[p]][i + 1, 4]) - (SP2[[p]][i, 4])
				SP2[[p]][, 10][is.na(SP2[[p]][, 10])] <- 0
				# creating values for dy
				SP2[[p]][i, 11] <- (SP2[[p]][i + 1, 5]) - (SP2[[p]][i, 5])
				# to remove NA and replace it with 0
				SP2[[p]][, 11][is.na(SP2[[p]][, 11])] <- 0
				# creating values for dis
				SP2[[p]][i, 12] <- sqrt(
					(SP2[[p]][i, 10]) ^ 2 + (SP2[[p]][i, 11]) ^ 2
				)
				# creating values for Square Speed
				SP2[[p]][i, 13] <- ((SP2[[p]][i, 12]) / TimeInterval) * 60
				return(SP2[[p]][i, c(10:13)])
			}
		)
	)
	SP2[[p]][1:MM, c(10:13)] <- as.data.frame(res)
	SP2[[p]][, c(10:13)] <- lapply(SP2[[p]][, c(10:13)], as.numeric)
	colnames(SP2[[p]]) <- c(
		"CellID", "CellName", "ImageID", "X", "Y", "R", "G", "B", "Phase",
		"dX", "dY", "Dis", "Speed"
	)
}

jump <- c()
for(i in 1:length(SP2)) {
	if (length(SP2[[i]][, 13]) < 144) {
		missing <- 144 - length(SP2[[i]][, 13])
		SP2[[i]][1:144, 13] <- c(rep(NA, missing), SP2[[i]][, 13])
	} else {
		extra <- 144 - length(SP2[[i]][, 13])
		SP2[[i]][1:144, 13] <- SP2[[i]][(1 + abs(extra)):length(SP2[[i]][, 13]), 13]
	}
	testB <- SP2[[i]][138:144, 13]
	testA <- SP2[[i]][130:137, 13]
	condition1 <- mean(testB[testB != 0])> mean(testA[testA != 0]) * 1.5
	if (condition1 & ((SP2[[i]][138, 13])>= 2.5 * mean(SP2[[i]][120:144, 13])) & ((SP2[[i]][138, 13])>= (3 * SP2[[i]][140, 13])| (SP2[[i]][138, 13])>= (3 * SP2[[i]][141, 13]) | (SP2[[i]][138, 13])>= (3 * SP2[[i]][142, 13])| (SP2[[i]][138, 13])>= (3 * SP2[[i]][139, 13]) | (SP2[[i]][138, 13])>= (3 * SP2[[i]][144, 13])| (SP2[[i]][138, 13])>= (3 * SP2[[i]][143, 13]))) {
		jump <- c(jump, i)}
	if (condition1 & ((SP2[[i]][139, 13])>= 2.5 * mean(SP2[[i]][120:144, 13])) & ((SP2[[i]][139, 13])>= (3 * SP2[[i]][140, 13])| (SP2[[i]][139, 13])>= (3 * SP2[[i]][141, 13]) | (SP2[[i]][139, 13])>= (3 * SP2[[i]][142, 13])| (SP2[[i]][139, 13])>= (3 * SP2[[i]][138, 13]) | (SP2[[i]][139, 13])>= (3 * SP2[[i]][144, 13])| (SP2[[i]][139, 13])>= (3 * SP2[[i]][143, 13]))) {
		jump <- c(jump, i)}
	if (condition1 & ((SP2[[i]][140, 13])>= 2.5 * mean(SP2[[i]][120:144, 13])) & ((SP2[[i]][140, 13])>= (3 * SP2[[i]][138, 13])| (SP2[[i]][140, 13])>= (3 * SP2[[i]][141, 13]) | (SP2[[i]][140, 13])>= (3 * SP2[[i]][142, 13])| (SP2[[i]][140, 13])>= (3 * SP2[[i]][139, 13]) | (SP2[[i]][140, 13])>= (3 * SP2[[i]][144, 13])| (SP2[[i]][140, 13])>= (3 * SP2[[i]][143, 13]))) {
		jump <- c(jump, i)}
	if (condition1 & ((SP2[[i]][141, 13])>= 2.5 * mean(SP2[[i]][120:144, 13])) & ((SP2[[i]][141, 13])>= (3 * SP2[[i]][140, 13])| (SP2[[i]][141, 13])>= (3 * SP2[[i]][138, 13]) | (SP2[[i]][141, 13])>= (3 * SP2[[i]][142, 13])| (SP2[[i]][141, 13])>= (3 * SP2[[i]][139, 13]) | (SP2[[i]][141, 13])>= (3 * SP2[[i]][144, 13])| (SP2[[i]][141, 13])>= (3 * SP2[[i]][143, 13]))) {
		jump <- c(jump, i)}
	if (condition1  & ((SP2[[i]][142, 13])>= 2.5 * mean(SP2[[i]][120:144, 13]))& ((SP2[[i]][142, 13])>= (3 * SP2[[i]][140, 13])| (SP2[[i]][142, 13])>= (3 * SP2[[i]][141, 13]) | (SP2[[i]][142, 13])>= (3 * SP2[[i]][138, 13])| (SP2[[i]][142, 13])>= (3 * SP2[[i]][139, 13]) | (SP2[[i]][142, 13])>= (3 * SP2[[i]][144, 13])| (SP2[[i]][142, 13])>= (3 * SP2[[i]][143, 13]))) {
		jump <- c(jump, i)}

	if (condition1 & ((SP2[[i]][143, 13])>= 2.5 * mean(SP2[[i]][120:144, 13])) & ((SP2[[i]][143, 13])>= (3 * SP2[[i]][140, 13])| (SP2[[i]][143, 13])>= (3 * SP2[[i]][141, 13]) | (SP2[[i]][143, 13])>= (3 * SP2[[i]][142, 13])| (SP2[[i]][143, 13])>= (3 * SP2[[i]][139, 13]) | (SP2[[i]][143, 13])>= (3 * SP2[[i]][144, 13])| (SP2[[i]][143, 13])>= (3 * SP2[[i]][138, 13]))) {
		jump <- c(jump, i)}
	if (condition1 & ((SP2[[i]][144, 13])>= 2.5 * mean(SP2[[i]][120:144, 13])) & ((SP2[[i]][144, 13])>= (3 * SP2[[i]][140, 13])| (SP2[[i]][144, 13])>= (3 * SP2[[i]][141, 13]) | (SP2[[i]][144, 13])>= (3 * SP2[[i]][142, 13])| (SP2[[i]][144, 13])>= (3 * SP2[[i]][143, 13]) | (SP2[[i]][144, 13])>= (3 * SP2[[i]][139, 13])| (SP2[[i]][144, 13])>= (3 * SP2[[i]][138, 13]))) {
		jump <- c(jump, i)}
	if (mean(testB[testB != 0])> 5 * (mean(testA[testA != 0]))) {
		jump <- c(jump, i)
	}
	jump <- jump[!duplicated(jump)]
}

length(jump) # number of cells with a jump
(length(jump) / length(SP2)) * 100 # % of cells with a jump
```

## Plotting tbe mean speed across the cell cycle phases for cells with a jump

```{r}
jumpDF <- data.frame()
for (j in jump) {
	jumpDF <- rbind(jumpDF, SP2[[j]][, 13])
}
dim(jumpDF)
nam <- c(1:144)
colnames(jumpDF) <- nam
for (i in 1:144) {
	jumpDF[length(jump) + 1, i]<- mean(jumpDF[1:length(jump), i], na.rm=TRUE)
	jumpDF[length(jump) + 2, i]<- sd(jumpDF[1:length(jump), i], na.rm=TRUE)
}
x <- time
y <- jumpDF[length(jump) + 1, ]
dy <- jumpDF[length(jump) + 2, ]
Time <- x
p <- graphics::plot(
	Time, y, type="l", col="red", xaxt="n", xlab="Cell Division Time (h)",
	ylab="Speed (um/h)", lwd=1, las=1, ylim=c(0, 220)
)
graphics::polygon(
	c(Time, rev(Time)), c(y + dy, rev(y-dy)), col="gray" , border=NA
)
graphics::lines(Time, y + dy, col="black")
graphics::lines(Time, y - dy, col="black")
graphics::lines(Time, y, type="l", col="red", lwd=3)
xx <- c(1, 12, 24, 36, 48, 60, 72, 84, 96, 108, 120, 132, 144)
yy <- c(-12, -11, -10, -9, -8, -7, -6, -5, -4, -3, -2, -1, 0)
axis(1, at=xx, labels=yy, col.axis="black", las=1, cex.axis=0.73)
```

## Plotting tbe mean speed across the cell cycle phases for cells without a jump

```{r}
SP2 <- split(Cells, Cells[, 2]) # splitting each cell based on phases
TimeInterval <- 5
for (p in 1:length(SP2)) {
	M <- SP2[[p]]
	MM <- length(M[, 1])
	res <- t(
		sapply(
			1:MM,
			function(i) {
				# creating values for dx
				SP2[[p]][i, 10] <- (SP2[[p]][i + 1, 4]) - (SP2[[p]][i, 4])
				SP2[[p]][, 10][is.na(SP2[[p]][, 10])] <- 0
				# creating values for dy
				SP2[[p]][i, 11] <- (SP2[[p]][i + 1, 5]) - (SP2[[p]][i, 5])
				# to remove NA and replace it with 0
				SP2[[p]][, 11][is.na(SP2[[p]][, 11])] <- 0
				# creating values for dis
				SP2[[p]][i, 12] <- sqrt(
					(SP2[[p]][i, 10]) ^ 2 + (SP2[[p]][i, 11]) ^ 2
				)
				# creating values for Square Speed
				SP2[[p]][i, 13] <- ((SP2[[p]][i, 12]) / TimeInterval) * 60
				return(SP2[[p]][i, c(10:13)])
			}
		)
	)
	SP2[[p]][1:MM, c(10:13)] <- as.data.frame(res)
	SP2[[p]][, c(10:13)] <- lapply(SP2[[p]][, c(10:13)], as.numeric)
	colnames(SP2[[p]]) <- c(
		"CellID", "CellName", "ImageID", "X", "Y", "R", "G", "B", "Phase",
		"dX", "dY", "Dis", "Speed"
	)
}

Total <- c(1:length(SP2))
noJump <- !is.element(Total, jump)
noJump <- Total[noJump]

for(i in 1:length(SP2)) {
	if (length(SP2[[i]][, 13]) < 144) {
		missing <- 144 - length(SP2[[i]][, 13])
		SP2[[i]][1:144, 13] <- c(rep(0, missing), SP2[[i]][, 13])
	} else {
		extra <- 144 - length(SP2[[i]][, 13])
		SP2[[i]][1:144, 13] <- SP2[[i]][(1 + abs(extra)):length(SP2[[i]][, 13]), 13]
	}
}

NOTjumpDF <- data.frame()
for (j in noJump) {
	NOTjumpDF <- rbind(NOTjumpDF, SP2[[j]][, 13])
}
dim(NOTjumpDF)
nam <- c(1:144)
colnames(NOTjumpDF) <- nam
for (i in 1:144) {
	NOTjumpDF[length(noJump) + 1, i] <- mean(
		NOTjumpDF[1:length(noJump), i], na.rm=TRUE
	)
	NOTjumpDF[length(noJump) + 2, i] <- sd(
		NOTjumpDF[1:length(noJump), i], na.rm=TRUE
	)
}
x <- time
y <- NOTjumpDF[length(noJump) + 1, ]
dy <- NOTjumpDF[length(noJump) + 2, ]
Time <- x

p <- graphics::plot(
	Time, y, type="l", col="red", xaxt="n", xlab="Cell Division Time (h)",
	ylab="Speed (um/h)", lwd=1, las=1, ylim=c(0, 220)
)
graphics::polygon(
	c(Time, rev(Time)), c(y + dy, rev(y - dy)), col="gray" , border=NA
)
graphics::lines(Time, y + dy, col="black")
graphics::lines(Time, y - dy, col="black")
graphics::lines(Time, y, type="l", col="red", lwd=3)
xx <- c(1, 12, 24, 36, 48, 60, 72, 84, 96, 108, 120, 132, 144)
yy <- c(-12, -11, -10, -9, -8, -7, -6, -5, -4, -3, -2, -1, 0)
axis(1, at=xx, labels=yy, col.axis="black", las=1, cex.axis=0.73)
```